Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-20 ms.tex
fixing the description of sigma

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy equation
accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}).  The first term approximates the
minimum amount of shock heating from stellar winds due to the random
movement of stars with respect to one another at the characteristic
velocity of $\sigma \approx \sqrt{GM_{\rm enc}/r}$.  This heating is
present even in old stellar populations for which mass and energy
injection are dominated by AGB winds with low velocities $\ll \sigma$.
The second term, $v_{w}^{2}$, parameterizes all additional sources of
heating, including faster main sequence or Wolf-Rayet stellar winds,
spin-down energy from millisecond pulsars, supernovae, AGN feedback,
etc.  As will be discussed in $\S\ref{sec:heating}$, the value of
$v_{w}$ will in general depend on the SMBH mass and the age of the
stellar population.  We further assume that $v_w$ is constant with
radius, i.e. that the volumetric heating rate is proportional to the
local stellar density.

>>>>

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy equation
accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}).  The first term approximates the
minimum amount of shock heating from stellar winds due to the random
movement of stars with respect to one another at the characteristic
velocity of $\sigma \approx \sqrt{\frac{3 G \Mbh}{(\Gamma+2)
    r}+\sigma_0^2}$, where $\sigma_0$ is a constant. The first term
corresponds to the velocity dispersion of a power law stellar density
profile with slope $-1-\Gamma$ in the presence of black hole. The
second term represents the observed constant velocity dispersion at
radii outside of the black hole sphere of influence \citep{ChaeBernardi+:2014a}. For a given
$\Mbh$, we take $\sigma_0^2\approx 3 \sigma_{\Mbh}^2$, where
$\sigma_{\Mbh}$ is the velocity dispersion from the
\citet{McConnellMa+:2011a} $\Mbh-\sigma$ relation.  This heating is
present even in old stellar populations for which mass and energy
injection are dominated by AGB winds with low velocities $\ll \sigma$.
The second term, $v_{w}^{2}$, parameterizes all additional sources of
heating, including faster main sequence or Wolf-Rayet stellar winds,
spin-down energy from millisecond pulsars, supernovae, AGN feedback,
etc.  As will be discussed in $\S\ref{sec:heating}$, the value of
$v_{w}$ will in general depend on the SMBH mass and the age of the
stellar population.  We further assume that $v_w$ is constant with
radius, i.e. that the volumetric heating rate is proportional to the
local stellar density.

### 2015-03-20 windheat.tex
fixing the description of sigma

% If we assume that a stellar population forms
% impulsively in the distant pass with IMF $\mu(m_*)$(with minimum mass
% $m_0$ and maximum mass $m_1$), then the surviving mass fraction at any
% future time $t$ is given by
% \begin{equation}
% f(t) =\frac{ \int_{m_0}^{m_{\rm T}(t)} m_* \mu(m_*) dm_* }{ \int_{m_0}^{m_1} m_* \mu(m_*) dm_* },
% \end{equation}
% where the main sequence turnoff mass is approximately
% \begin{equation}
% m_{\rm T}(t) \approx 2.5M_\odot~ \left( \frac{t}{10^9~{\rm yr}} \right)^{-0.4}.
% \end{equation}
% For a Salpeter IMF $\mu(m_*) \propto m_*^{-2.35}$ with $m_0=0.1M_\odot$ and $m_1=100M_\odot$,
% \begin{equation}
% f_{\rm Sal}(t) = 1.098 - 0.490 \left(\frac{t}{10^{10}~{\rm yr}} \right)^{0.14}
% \end{equation}
% {\bf NCS: we should probably use a Kroupa/Chabrier IMF, but this gets the ball rolling.}
% If we approximate post-main sequence evolution as instantaneous and define $\lambda(\Mstar)$ as the fractional mass lost during all stages of stellar evolution, then the mass loss rate density
% \begin{equation}
% q(t) = \frac{\rho_*}{\bar{m}_*} \lambda(m_{\rm T}(t)) m_{\rm T}(t) \frac{df}{dt},
% \end{equation}
% where the mean stellar mass $\bar{m}_* = \int_{m_0}^{m_{\rm T}(t)} \Mstar\mu(\Mstar)d\Mstar \approx 0.3 M_\odot$.  Further approximating $\lambda(\Mstar)=0.5$, and using the Salpeter IMF once more, gives
% \begin{equation}
% q(t) = \frac{\rho_*}{10^{10}~{\rm yr}} \times 0.11 \left(\frac{t}{10^{10}~{\rm yr}} \right)^{-1.26}.
% \end{equation}
% This is a specific, time-dependent definition of $\eta(t) (=0.11(t/t_{\rm h})^{-1.26})$; if we consider different star formation scenarios (for example, continuous star formation) or different IMFs, it will change.  Once these free parameters are specified, however, we can answer an important question: do young stellar populations increase or decrease the SMBH feeding rate $\dot{M}$?  Clearly, $\eta(t)$ is larger for young stellar populations, but these stars also have high wind velocities that diminish the stagnation radius.  Crudely approximating $v_{\rm w}=75~{\rm km~s}^{-1}$ for $m_{\rm T} < 10M_{\odot}$ and $v_{\rm w}=3000~{\rm km~s}^{-1}$ for $m_{\rm T} > 10M_{\odot}$ (motivated by the transition from dust-driven wind loss on the AGB to line-driven wind loss from Wolf-Rayet stars), we can employ the relation $\dot{M} \propto \eta(t) r_{\rm s}^{2-\Gamma}\propto \eta(t) v_{\rm w}^{-4+2\Gamma}$ (where $\rho_* \propto r^{-\Gamma}$) to determine the impact of stellar ``youth'' on SMBH feeding rates.
% {\bf AG:As I previously mentioned this discussion of the eta is not
%   quite correct...}
% \begin{figure}
% \includegraphics[width=\columnwidth]{NickPlot.eps}
% \caption{\label{NickPlot} SMBH feeding rates $\dot{M}=\eta(t) \times \Mstar(r_{\rm s})$, in arbitrary units.  The green, blue, and black curves are for galaxies with $\Gamma$ values of $0.1$, $0.5$, and $0.9$, respectively.  Solid curves represent impulsive-mode star formation, while dotted curves represent continuous-mode star formation. {\bf NCS: I think these old continuous curves are wrong, need to revise}}
% \end{figure}

% In Fig. \ref{NickPlot} we plot $\dot{M}$, in arbitrary units, as a
% function of time, for three different stellar density profiles $\Gamma
% = \{1.1, 1.5, 1.9\}$ (which are normalized to have the same mass at an
% influence radius $r _{\rm soi}=10~{\rm pc}$ around a $10^7M_\odot$
% SMBH).  We parametrize the wind velocity as
% \begin{equation} \frac{v_{\rm w}}{3000 ~\rm
% km~s^{-1}}=520-495\tanh\left( \frac{t-10^{7.5}~{\rm yr}}{10^{7}~{\rm
% yr}}\right). \label{NickV1}
% \end{equation} This counts Type II SNe heating as ``winds;'' if
% instead we are in the portion of parameter space where $r_{\rm II}$ is
% very large, then we use the alternate parametrization
% \begin{equation} \frac{v_{\rm w}}{3000 ~\rm
% km~s^{-1}}=520-495\tanh\left( \frac{t-10^{7}~{\rm yr}}{10^{6.5}~{\rm
% yr}}\right), \label{NickV2}
% \end{equation} which only allows short lived Wolf-Rayet stars to
% contribute to the high-heating mode.  The ``impulsive burst'' mode of
% star formation produces large ($\sim 10$) differences between the
% three $\dot{M}$ curves at early times, when $r_{\rm s}$ is small, but
% smaller ($\sim 3$) differences at late times, when $r_{\rm s}$ is
% large.  We also plot, as dotted curves, a simple model for the
% ``continuous'' mode of star formation, where mass loss is calculated
% as $\bar{\eta} = \int\eta(t)dt/t_{\rm h} \approx 4$ and an average
% energy injection in the wind is calculated as $\bar{v_{\rm
% w}^2}=\int\eta(t)v_{\rm w}^2(t)dt/\bar{\eta} \approx (800~{\rm
% km~s}^{-1})^2$.  Interestingly, the continuous mode of star formation
% produces small differences from late-time $\dot{M}$ seen in cusp
% galaxies with impulsive star formation; however, continuous mode star
% formation decreases late-time $\dot{M}$ by an order of magnitude
% relative to impulsive star formation in core galaxies.  {\bf NCS: I
% think this old discussion of MDot in the continuous limit is wrong,
% need to revise}

@@

The MS wind mass loss rate $\dot{m}(\Mstar, t)$ is calculated based on the generalization of Reimer's law given in \citet{SchroderCuntz:2005a} (their eq.~[4]) as
\begin{align}
  \dot{m}=8 \times 10^{-14} \frac{L_* R_*}{\Mstar}
  \left(\frac{T_{\rm eff}}{\rm 4000 K}\right)^{3.5}
  \left(1+\frac{g_{\odot}}{4300 g_*}\right) \Msun \pyear,\
\end{align}
where  $R_*$, $L_*$, $T_{\rm eff}$ and $g_*$ are the stellar radius,
luminosity, effective temperature, and surface gravity, respectively, the latter normalized to its solar value $g_{\odot}$.  The stellar radius and luminosity are estimated as (\citet{Kippenhahn&Weigert90}; Figs.~22.2) 22.3)
\begin{align}
L_*=
\begin{cases}
L_{\odot} (\Mstar/\Msun)^{3.2} & \Mstar > \Msun \\
L_{\odot} (\Mstar/\Msun)^{2.5} & \Mstar \le \Msun
\end{cases}
\end{align}
\begin{align}
R_*=
\begin{cases}
R_{\odot} (\Mstar/\Msun)^{0.57} & \Mstar > \Msun \\
R_{\odot} (\Mstar/\Msun)^{0.8} & \Mstar \le \Msun
\end{cases}
\end{align}
The wind velocity of main sequence winds is assumed
to equal $v_w (\Mstar, t) = v_{w,\odot}(M_{\star}/M_{\odot})^{1/2}(R_{\star}/R_{\odot})^{-1/2}$, i.e.~scaling as the stellar escape velocity and normalized to the velocity of the solar wind $v_{w,\odot} = 430$ km s$^{-1}$; this produces an effective wind heating velocity for main
sequence winds alone of $\sim 100$ km s$^{-1}$ for $\tau_{\star} \sim t_{\rm h}$, close to the value found by \citet{NaimanSoares-Furtado+:2013a} for globular clusters based on a more
sophisticated population synthesis treatment.

>>>>

@@

{\bf AG: Go back to the old mass loss prescription ASAP so that we don't
  violate mass conservation!}
The MS wind mass loss rate $\dot{m}(\Mstar, t)$ is calculated based on
the generalization of Reimer's law given in
\citet{SchroderCuntz:2005a} (their eq.~[4]) as
\begin{align}
  \dot{m}=8 \times 10^{-14} \frac{L_* R_*}{\Mstar}
  \left(\frac{T_{\rm eff}}{\rm 4000 K}\right)^{3.5}
  \left(1+\frac{g_{\odot}}{4300 g_*}\right) \Msun \pyear,\
\end{align}
where  $R_*$, $L_*$, $T_{\rm eff}$ and $g_*$ are the stellar radius,
luminosity, effective temperature, and surface gravity, respectively, the latter normalized to its solar value $g_{\odot}$.  The stellar radius and luminosity are estimated as (\citet{Kippenhahn&Weigert90}; Figs.~22.2) 22.3)
\begin{align}
L_*=
\begin{cases}
L_{\odot} (\Mstar/\Msun)^{3.2} & \Mstar > \Msun \\
L_{\odot} (\Mstar/\Msun)^{2.5} & \Mstar \le \Msun
\end{cases}
\end{align}
\begin{align}
R_*=
\begin{cases}
R_{\odot} (\Mstar/\Msun)^{0.57} & \Mstar > \Msun \\
R_{\odot} (\Mstar/\Msun)^{0.8} & \Mstar \le \Msun
\end{cases}
\end{align}
The wind velocity of main sequence winds is assumed
to equal $v_w (\Mstar, t) = v_{w,\odot}(M_{\star}/M_{\odot})^{1/2}(R_{\star}/R_{\odot})^{-1/2}$, i.e.~scaling as the stellar escape velocity and normalized to the velocity of the solar wind $v_{w,\odot} = 430$ km s$^{-1}$; this produces an effective wind heating velocity for main
sequence winds alone of $\sim 100$ km s$^{-1}$ for $\tau_{\star} \sim t_{\rm h}$, close to the value found by \citet{NaimanSoares-Furtado+:2013a} for globular clusters based on a more
sophisticated population synthesis treatment.
